Working title: Ecological network control

### 2018-06-15 paper/manuscript.Rmd
Amend Introduction
Close #35

A major goal in ecology is to understand the roles played by different species in the biotic environment.
Within community ecology, a complex-systems approach has led to the development of a variety analytical and simulation tools with which to compare and contrast the roles of species embedded in a network of interactions [@bascompte_assembly_2009; @stouffer_evolutionary_2012].
<!-- Add some older references as this is a long-standing research line; see the 2012 paper for some if you'd like. -->
A particularly relevant dimension of any species' role is its ability to alter the abundance of other species in the community—since changes of this nature impact ecosystem function and biodiversity, and might have knock-on effects on other processes and services.
This ability is often referred to as a species' "keystoneness" [refmissing].

A significant proportion of the network tools used to estimate species roles in this context rely on the assumption that there is a link between a given species' centrality—a relative ranking of its importance that stems from social network research—and its ability to alter the abundances of other species. 
Taken at face value, the basis of this assumption is reasonably clear: a major justification for representing ecological communities as networks is the idea that the abundance of a species depends on the abundance of its neighbours (those species with which it interacts) and, by extension, the abundance of its neighbours' neighbours, and so on. 
By definition, more central species are better connected and hence should be more likely to affect the abundance of other species in the network.

<!-- Beware that the paragraph above could rub some readers/reviewers (those that believe in the usefulness of centrality) the wrong way. -->

<!-- I believe it should be 'keystoneness' and not 'keystoness'. I've tried to correct all, but please check this. -->

<!-- I didn't do it, but I would think about joining the text below (up to "more mechanistically-grounded") with the above to make a single paragraph. It'd be long, but the thoughts flow well. -->

Though conceptually intuitive, the relationship between centrality and keystoneness is largely phenomenological. 
On one hand, there is more than a dozen indices of centrality used in ecology. 
Choosing a particular metric can be a difficult task, especially because to date there are no clear methods to determine the suitability of an index to a particular application.
<!-- I feel like the question of there being so many centralities distracts from the main line of thinking. You could arguably emphasize the phenomenological piece better (with an example?) and jump straight to the sentence below this comment. -->
Furthermore, recent studies have shown that definitions of keystoneness based on centrality might not adequately predict the cascading extinctions keystone species are usually associated to. 
As a result, community ecology could arguably benefit from an alternative, more mechanistically-grounded, approach to understand how species affect each other's abundance.

Species abundances are also influenced by the specific dynamic relationships between nodes and the mechanisms of self-regulation; however, community and population dynamics can be modelled in innumerable ways, and empirical support for one versus another is often still ambiguous. 
The alternative approach should therefore acknowledge ecosystem dynamics, but without being overly dependent on the particular choices of how they are characterised.
Among the various possibilities, *control theory* appears to be a strong candidate [@isbell_human_2013]. 
Control theory is a widely-studied branch of engineering used to determine and supervise the behaviour of dynamical systems [@motter_networkcontrology_2015]. 
It is inherently designed to deal with system feedbacks and its application has recently been expanded to complex networks [@liu_control_2016]. 
In concordance with long-standing ecological questions, advances in this field have established a clear link between the structure of the network and the way nodes affect each other. 
Unlike centrality indices, however, this link is not based on a priori assumptions between network metrics and keystoneness but is instead based on well-established advances in both dynamical and complex-systems theory.
<!--STOUFFER: Are we shooting ourselves in the foot since this isn't entirely true? That is, see comments from ELE about linearity and otherwise. CAGUA: How about this modified phrasing? DBS: I approve. ;)-->.

Despite the apparent overlap, the concepts and approaches of control theory are not yet commonly applied in an ecological context. 
Possibly, this is in part because research on the control of complex networks is generally more concerned with the state of the system as a whole. 
This is, in contrast with the interest in ecological roles, where the focus is on the nodes that compose the network.
At its fundamental level, control theory first determines whether a system is controllable or not; that is, it asks if a system can be driven to a desired state within a finite amount of time. 
Although the controllability of a network is a whole-system property, it has been recently shown that asking for the controllability of a complex-system is equivalent to finding a particular set of nodes: the set with which is possible to control the state of the whole network.
Notably, this set of nodes is not always unique for a given network.
This implies that an examination of the distinct sets provides a means to connect nodes with its *general* ability to modify the system to which they belong.

<!-- I feel like the paragraph above is trying to tell me two things and does neither very clearly. My gut tells me that it could probably be improved by removing the first sentence and worrying less about why it hasn't but ways in which the application to ecology could be improved. Does that make sense? -->

Here, we apply methods from control theory to a particular ecological problem and show how it can be used to gain insight into the role of species in an ecological network. 
Specifically, we outline the approach using a set of ten pairs of uninvaded and invaded plant-pollinator communities. 
We use invaded communities because there is strong empirical evidence showing that invasive species play an important role shaping the abundances of other species, something which is particularly true in these ten networks [@lopezaraizamikel_impact_2007; @bartomeus_contrasting_2008-1]. 
This choice thus offers us an opportunity to explicitly contrast our theoretical observations with empirical evidence.
With these network, we first ask whether there are differences between the controllability of invaded and uninvaded networks. This question is motivated by in the difficulties commonly encountered with invasive-species eradication and ecosystem restoration [@woodford_confronting_2016].
<!--STOUFFER: I'd remove the grounded in.... stuff, but that's just me.  CAGUA: I don't like it either. I tried removing but then it feels like this question is coming out of the blue.-->
We then expand existing methods from control theory to effectively link the *controllability* of a network with the role of particular species.
We ask—from a control-theoretic perspective—whether there are key differences between species in the role they play at driving the population of other members of the community and identify the ecological factors related to these differences. 
Finally, we compare the proposed approach to current methods based on species centrality and show how these methods are indeed valuable but ultimately paint a limited picture in regards to the dynamic "keystoneness" of a species.

>>>>

A major goal in ecology is to understand the roles played by different species in the biotic environment.
Within community ecology, a complex-systems approach has led to the development of a variety analytical and simulation tools with which to compare and contrast the roles of species embedded in a network of interactions [@guimera_cartography_2005; @bascompte_assembly_2009; @stouffer_evolutionary_2012; @coux_linking_2016].
A particularly relevant dimension of any species' role is its ability to alter the abundance of other species in the community—since changes of this nature impact ecosystem function and biodiversity, and might have knock-on effects on other processes and services.
This ability is often referred to as a species' "keystoneness" [@mills_keystone-species_1993].

A significant proportion of the network tools used to estimate species roles in this context rely on the calculation of a species' centrality—a relative ranking of its importance that stems from social network research [@noah_e._friedkin_theoretical_1991].
Generally speaking, central species tend to be present in more food chains, are better connected, and have broader niches.
Centrality metrics have been shown to be useful tools to rank species in regards to their risk of extinction or the potential secondary extinctions after the species is lost [@dunne_network_2002; @kaiser-bunbury_robustness_2010]. 
Although there are some differences among centrality metrics—as each of them emphasizes a different aspect of a species' interaction patterns—they all tend to be correlated with the way species participate in the network's food chains [@jordan_quantifying_2007]. 
Because of this property, centrality metrics have been often used to identify keystone species.

This relationship between centrality and keystoneness relies on an important assumption. 
That that there is a link between a given species' centrality and its ability to alter the abundances of other species. 
The basis of this assumption is reasonably clear: a major justification for representing ecological communities as networks is the idea that the abundance of a species depends on the abundance of its neighbours (those species with which it interacts) and, by extension, the abundance of its neighbours' neighbours, and so on. 
By definition, more central species are better connected and hence should be more likely to affect the abundance of other species in the network.

Despite being conceptually intuitive, the relationship between centrality and keystoneness is largely phenomenological. 
<!-- Daniel: I'm struggling to find an example of this idea. Does anything come to your mind? -->
Often, dramatic changes in ecosystem functioning can also occur without complete removal of a species. 
Hence, we expect the usefulness of centrality to diminish as we move beyond species removal towards understanding how species relate to the state and stability of the ecosystem. 
Particularly so, when the state of an ecosystem is underpinned by more than a single species. 
As a result, community ecology could arguably benefit from an alternative more mechanistically-grounded approach to understand how species affect each other's abundance.

Species abundances (and consequently the state of the community) are influenced by both the species interactions and the specific dynamic relationships between nodes and the mechanisms of self-regulation. 
However, community and population dynamics can be modelled in innumerable ways, and empirical support for one versus another is often still ambiguous. 
The alternative approach should, therefore, acknowledge ecosystem dynamics, but without being overly dependent on the particular choices of how they are characterised.
Among the various possibilities, *control theory* appears to be a strong candidate [@isbell_human_2013]. 
Control theory is a widely-studied branch of engineering used to determine and supervise the behaviour of dynamical systems [@motter_networkcontrology_2015]. 
It is inherently designed to deal with system feedbacks and its application has recently been expanded to complex networks [@liu_control_2016]. 
In concordance with long-standing ecological questions, advances in this field have established a clear link between the structure of the network and the way nodes affect each other. 
Unlike centrality indices, however, this link is not based on a priori assumptions between network metrics and keystoneness but is instead based on well-established advances in both dynamical and complex-systems theory.

At its fundamental level, control theory first determines whether a system is controllable or not; that is, it asks if a system can be driven to a desired state within a finite amount of time. 
Although the controllability of a network is a whole-system property, it has been recently shown that asking for the controllability of a complex-system is equivalent to finding a particular set of nodes: the set with which is possible to control the state of the whole network.
Notably, control theory is consistent with the observation that to change the state of a network, more than a ranking, we need a set of nodes.
Importantly, this set of nodes is not always unique for a given network.
This implies that an examination of the distinct sets provides a means to connect nodes with its *general* ability to modify the system to which they belong.

Here, we apply methods from control theory to a particular ecological problem and show how it can be used to gain insight into the role of species in an ecological network. 
Specifically, we outline the approach using a set of ten pairs of uninvaded and invaded plant-pollinator communities. 
We use invaded communities because there is strong empirical evidence showing that invasive species play an important role shaping the abundances of other species, something which is particularly true in these ten networks [@lopezaraizamikel_impact_2007; @bartomeus_contrasting_2008-1]. 
This choice thus offers us an opportunity to explicitly contrast our theoretical observations with empirical evidence.
With these networks, we first ask whether there are differences between the controllability of invaded and uninvaded networks. 
This question is motivated by the difficulties commonly encountered with invasive-species eradication and ecosystem restoration [@woodford_confronting_2016].
<!--STOUFFER: I'd remove the grounded in.... stuff, but that's just me.  CAGUA: I don't like it either. I tried removing but then it feels like this question is coming out of the blue.-->
We then expand existing methods from control theory to effectively link the *controllability* of a network with the role of particular species.
We ask—from a control-theoretic perspective—whether there are key differences between species in the role they play at driving the population of other members of the community and identify the ecological factors related to these differences. 
This allows us to identify species that are critical to changes in the ecosystem state and show that they have a larger than expected impact on the structural stability of the community. 
Finally, we compare the proposed approach to current methods based on species centrality and show how these methods are indeed valuable but ultimately paint a limited picture in regards to the dynamic "keystoneness" of a species.
